\chapter{Background and Related Work}

\section{Affective Computing and Facial Emotion Recognition}

Affective computing is a research area focused on systems that can recognize, interpret, and react to human emotions. The field was introduced by Picard, who argued that emotion is an important part of intelligent human-computer interaction \cite{picard1997affective}.

In real communication, people do not use only words. They also use facial expressions, voice tone, gestures, and context. Because of this, emotion-aware systems may analyze several types of input, such as face images, speech, and text \cite{poria2017review,mobbs2025emotionreview}. This dissertation focuses mainly on facial emotion recognition, while speech is used only as a way of converting the user's voice into text.

Facial Emotion Recognition is the task of predicting a person's apparent emotion from an image or video frame of the face. Common emotion classes include anger, disgust, fear, happiness, sadness, surprise, and neutrality. In my project, the detected facial emotion is not treated as a certain description of the user's real emotional state. Instead, it is used as additional context for the assistant.

This distinction is important because facial emotion recognition is a difficult task. In real world webcam conditions, predictions can be affected by lighting, camera quality, face position, occlusions, and the ambiguity of facial expressions. Recent surveys also show that performance depends strongly on the dataset, the model architecture, and the evaluation setting \cite{mobbs2025emotionreview}. For this reason, the assistant uses the predicted emotion as a soft signal, not as a rule that controls the direction of generated answer.

\section{Public Datasets and Benchmarks for Facial Emotion Recognition}

Facial emotion recognition models are usually trained and compared on public datasets. One of the most used benchmarks is FER2013, introduced in the Challenges in Representation Learning competition \cite{goodfellow2015challenges}. FER2013 contains low-resolution facial images labeled with seven emotion classes: angry, disgust, fear, happy, sad, surprise, and neutral.

FER2013 is public and widely used, which makes it easier to compare the results with other projects. However, the dataset also comes with some limitations. The images have low resolution, some labels can be noisy, and the emotion classes are not perfectly balanced. These aspects can affect both the training process and the final evaluation.

Other datasets also exist for facial emotion recognition. AffectNet is a larger dataset that contains both categorical emotion labels and valence-arousal annotations \cite{mollahosseini2019affectnet}. Other public examples are CK+, JAFFE, RAF-DB, and FERPlus. These datasets differ in size, image quality, annotation method, and acquisition conditions.

Recent work still uses FER2013 to evaluate modern architectures. For example, Naik et al. use an EfficientNetB2-based model and report results on FER2013, while also pointing out problems such as low image quality, label noise, and class imbalance \cite{naik2026efficientnetb2}. This makes FER2013 a relevant choice for the solution presented in the current thesis, especially because the goal is not only classification accuracy, but also integration into a real-time assistant.

\section{Existing Components for Local AI Assistants}

Open-source tools allow AI assistant components to run directly on the user's computer. This is useful because the application may handle personal information, such as messages, voice recordings, or webcam images.

For speech-to-text, one of the most widely used open-source options is Whisper, a pretrained model for automatic speech recognition \cite{radford2022whisper}. Other available alternatives are wav2vec 2.0, which is based on self-supervised speech representation learning, and Vosk, which provides offline speech recognition models \cite{baevski2020wav2vec}.

For response generation, large language model families provide versions that can be executed locally. Examples include Llama, Gemma, Mistral, Qwen, and Phi \cite{touvron2023llama2,gemma2024,jiang2023mistral}. Tools such as Ollama simplify local deployment by downloading models, exposing them through a local API, and handling inference on the user's machine.

PyTorch and torchivision are the most used frameworks for training deep neural networks for computer vision tasks. In this combination, OpenCV is a great additional framework that helps with webcam access, face detection, frame processing, and image manipulation with high quality and performance. These tools represents the basic requirements for applications that combine vision, speech and language on the same device.